### 2019-06-04 chapter_00_open_science.tex
Revise chapter opensci

In order to attain the research goals in this thesis, a path off the beaten
track was pursued -- i.e. to create a handful of easy to maintain, rigorously
tested, and most-importantly, open-source scientific packages which does not
compromise on performance.
%
This resulted in the creation of FluidDyn, a project to facilitate open-science
in the field of fluid mechanics.
%
The focus of the project was to develop a set of packages to implement a
framework for simulations \citep{fluidsim}, experiments
\citep{augier_fluidimage_2016,augier_fluidlab_nodate}, Fast Fourier Transforms
\citep{fluidfft} -- all depending on a base package to reuse certain utilities
and data structures \citep{fluiddyn}.
%
The project owes a lot to the recent developments in methods and tools for
open-source software engineering.
%
Several good and modern coding practices were adopted while developing these
packages.
%
An understanding of the design choices and the workflow involved would be of
general interest to the reader.

FluidDyn packages are primarily written in Python language. Incidentally, it is
the defining characteristics of Python which make the eventual success of such
a project realizable.
%
Python is one of the most important tool in recent open-source dynamics,
particularly in science.
%
The popularity of the language in fluid mechanics is growing, but is yet
to reach widespread adoption as the language of choice.

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline, have flourished a lot in the latter half of the twentieth century.
Incidentally the origin of this discipline can be traced back to the ideas in
\citet{richardson_weather_1922} to achieve numerical weather prediction.
Following the arrival of first electronic general-purpose computer, ENIAC, the
first simulations were made using a simple barotropic model to make 24 hour
forecasts \citep{lynch_richardson_2010}. In those early days when the
availability of computers was limited, such methods were neither widespread nor
feasible. Scientific investigations were dominated by theoretical and
experimental methods.
%
We have come a long way since then and today computational methods complement
theoretical and experimental studies, and are equally important.
%
% Thus, in a typical scientific study, there is significant effort involved in
% translating the mathematical representation to a working research software.
%
With the connectivity provided by the world wide web it is possible to achieve
more and reinvent the way we do sciences.

@@

In recent years, there has been a move towards open science
\citep[see the report by][]{royal_society_great_britain_science_2012}.
%
Depicted in figure~\ref{fig:opensci} are some of steps needed to ensure
reproducibility and transparency in sciences.
%
Traditionally, \emph{knowledge} has been disseminated in the form of courses
and books from universities, and to the wider scientific community through
research articles, conference proceedings and workshops.
%
In the last decade, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals\footnote{In the European
  Union, a proposal to implement Plan S (\url{https://www.coalition-s.org}) was
  been set in motion in September 2018 to fast track adoption of open-access.
} either through systematic
reforms in academia on how researchers' productivity are assessed or through
regulations and offering incentives, such as waiving the application processing
charge for publishing \citep{nosek_promoting_2015}.

A lot of work is required to condense and implement this knowledge in the form
of working applications.
%
The \emph{know-how} can be made accessible and open to scrutiny by making
research software open-source with an appropriate license\footnote{For a
	detailed comparison of licenses, visit either
	\url{https://choosealicense.com} or
	\url{https://www.gnu.org/licenses/license-list.html}. }.
%
To make codes usable they have to be complemented by documentation which
typically includes tutorials, examples and a detailed commentary of different
components.
%
Furthermore, to inspire confidence in and to ensure reliability of a code
throughout the process of software development, it needs unit-tests through
which a code is continuously built, tested and deployed --- a process known as
Continuous Integration (CI). Depending on how matured a software is, CI can be
easy or cumbersome to implement, but nevertheless, the result is a cleaner and
reliable code.

Finally, the \emph{results} generated from the such codes together with the
workflow (i.e.\ scripts to run the codes and to post-process the data) when made
open in the form of datasets can ensure that research is reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

Open-source software development that we know today has been built on top of
the firm foundations of the Free and Open-Source Software (FOSS) movement. The
term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success for scientific publication
of the free type-setting standards \TeX\ (1977) and \LaTeX\ (1985).
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development, as described in \citet{raymond_cathedral_1999}, wherein a
code can stay structured despite releasing frequently and delegating tasks to
community with myriad agendas.

@@

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial).
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
In web-based source development platforms like GitHub, Bitbucket and GitLab,
DVCS repositories are enhanced with code-review tools, typically consisting of
a issue tracker and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and lowering the learning
barrier for scientists who are not experts in software development is crucial
for our project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

\subsubsection{Continuous Integration and documentation} The need for
strong reliability of packages with increasing complexity has forced developers
to use Continuous Integration (CI).
%
When the code depends on a wide variety of external dependencies, CI is useful
to avoid \fnref{https://en.wikipedia.org/wiki/Software\_rot}{software rot},
i.e.\ to avoid obsolescence with newer versions of the compiler or its
dependencies. CI is specifically important for dynamical languages and
open-source code since, it allows open-source projects to welcome contributions
from other developers while maintaining a high degree of reliability.
%
Firstly the code is \emph{built} against a known, repeatable emulation
environment on a server. Thereafter, a set of \emph{unit tests} are run for
each commit (modification of the code) to check that no bugs have been
introduced. It is also important to analyse what portion of the source code is
covered by these unit tests --- a method referred to as \emph{code coverage}.
%
Almost all CI platforms are free for open-source projects --- Travis CI,
CircleCI and Appveyor to name a few. Bitbucket and GitLab have built-in CI
implementations. Websites such as Codecov and Coveralls help to chart and
analyse the code coverage history, viewable by anyone. To use these services,
one simply has to write a simple file in YAML format and authenticate the
respective services with ``read'' permissions.

New users and developers often rely on good documentation and examples to start
working with an unfamiliar piece of software. Documentation generators such as
Sphinx (for Python) and Doxygen (for many languages including C, C++, Fortran
and Python) parses in-line comments into documentations as PDF, HTML and more
\citep{lee_ten_2018}. The documentation generation can be automated online and
deployed using any CI. ``Read the Docs'' has, in the recent years, become the
hub for building and hosting documentation online, thus facilitating the
process for developers and considerably improving access to information for
users.
%
In the FluidDyn project we this stop at automating tests and documentation of
the latest version, but it is possible to further exploit the CI by adding code
style checks by running ``linters'' alongside unit tests, \emph{continuous
delivery} of package releases and preserving older versions of documentation.

@@

% Mutual aid and knowledge sharing have been important aspects of the web since
% its inception and these practices are now of foremost importance for many
% human activities.
%
%In particular, 
Learning programming and computer science is now greatly aided by websites like
the community driven encyclopaedia \href{https://www.wikipedia.org/}{Wikipedia}
or the community driven forum \href{https://stackoverflow.com/}{Stack
Overflow}, where people write thousands of questions and answers on programming
each week. IRC protocol has continued to sustain the test of time. Instant
messaging channels such as \codeinline{\#python} on Freenode facilitate
interactive discussions and a form of community-level support system --- a role
which used to be filled by mailing lists alone in the past. Alternatives such
as Riot (Matrix protocol), Gitter, and Slack have gained popularity offering
different integration solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as in the case of
the FluidDyn project, can achieve near real-time collaboration without having
to arrange video conferences or meetings.

Thus, there is a number of new methods and tools in software development.
Taken all together, they greatly improve our collective efficiency and
open possibilities of interactions that were unthinkable a few years ago.
%
For example, the success of community-driven software like \pack{astropy} (a
subject-area research library with
\href{https://github.com/astropy/astropy}{more than 240 contributors}) and
\pack{scikit-learn} (a machine learning toolkit with
\href{https://github.com/scikit-learn/scikit-learn}{more than 1000
contributors}) would not have been possible without these new software
development methods.
%
As scientists, we can wonder how to fully exploit these new possibilities to do
science.  The project FluidDyn is an attempt to help the fluid dynamics
community to advance in this direction.

@@

The Python community has thrived remarkably in the last few years, which is
evident from the fact that by the end of 2017, Python became the second most
popular language in terms of pull-requests in GitHub\footnote{See GitHub's
	annual report titled \href{https://octoverse.github.com/}{ ``The State
of the Octoverse 2017''}.} and questions on StackOverflow\footnote{See
\href{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}{ ``The
Incredible Growth of Python''}, a blog post by Stack Overflow.}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{See
	the analysis titled
	\href{https://spectrum.ieee.org/computing/software/%
the-2017-top-programming-languages}{``The 2017 Top Programming Languages''} on
IEEE spectrum.} ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings agree that Python usage has been steadily growing and is here to stay.
Python largely outclasses languages like Fortran and Matlab according to these
statistics, and is now clearly one of the mainstream languages with C, Java,
C++ and JavaScript.
%
The language has been designed to boost the communication of technical ideas
between humans. 
%
Let us summarize some characteristics of the Python languages that have lead to
such an incredible success.
%
Python differs from other mainstream languages for a variety of reasons:

@@

\item \textbf{Batteries included.} Python comes with a large and high quality
\href{https://docs.python.org% /3/library/index.html}{standard library}.

@@

\end{itemize}
%
Thus, Python is one of the best ``glue'' languages for fast prototyping.
Code development in Python is much faster and easier than with many other
languages. Less bugs are introduced by the developers simply because there are
fewer lines of code \citep{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also facilitated by the very good code
readability associated with the language.
%
Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy from the start and the
complexity gradually increases as the level of the developer increases.
%
Its interpreted nature is advantageous for a rapid development cycle and the
possibility of interactive workflow (usually with IPython or Jupyter
notebooks).
%
Therefore, Python is good for developers of all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
%
Note that there are many implementations of Python
interpreters\footnote{Notable ones are CPython (written in C), Jython (Java),
IronPython (C\#), PyPy (Rpython, a subset of Python) and
MicroPython/CircuitPython (C, targeted to micro-controllers).}. The default and
most widely used implementation is written in C and is called CPython.
%
A hard and controversial decision was taken to clean up Python from its
historical incoherences by having a new series of backwards-incompatible
versions of the language, starting with Python 3.0, released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
behind us for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible as of 2018.}. Especially for scientific applications,
we can now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
releases of \Numpy} and of \href{http://www.python3statement.org/}{the main
scientific packages} will soon drop compatibility with Python 2. These changes
open doors to very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science any more and
instead, adopt the newer versions.
%
New versions of Python have nice new features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).

@@

\end{itemize}
%
Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established
foundational packages
%
(\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{https://pyzo.org}{Pyzo},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

Python has a very large and supportive community (see, for example,
\fnref{http://stackoverflow.com/tags/python}{Stack Overflow tags}).
%
Python development is also supported by companies, and most of them specialize
in web technologies and data-driven research. However, recently, companies
involved in scientific software development have started to become quite
influential. For example, Anaconda Inc.\ is strongly engaged in the development
of tools like JIT Python accelerator \pack{Numba}, and parallelism framework
\pack{dask}. Hardware giants such as Intel and Nvidia also contribute to the
Python scientific ecosystem.
%
To summarize, it is practical to use and extend Python for scientific purposes.

@@

\item \textbf{Concurrent, but not parallel multi-threading.} Threads behave
	like light sub-programs. In principle, they can use the different cores
	of the CPU at the same time, but the CPython interpreter forbid threads
	to interpret python code at the same time\footnote{See
	\href{https://opensource.com/article/17/4/grok-gil}{this blog article
on ``How to grok the GIL''} and
\href{https://faster-cpython.readthedocs.io/cpython37.html}{this technical
presentation by a CPython core developer}.}. In other words, threads can
coexist concurrently, by not execute in parallel.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
GIL greatly simplifies the implementation of CPython and that
it is very difficult to remove it while keeping other nice technical properties
of CPython intact\footnote{See
\url{https://wiki.python.org/moin/GlobalInterpreterLock}. Note that PyPy
also uses a GIL, while two other implementations of the Python language, Jython
and IronPython, do not have this limitation}.

\item \textbf{Lively, and thus, complicated ecosystem of packages.} For many
	applications, the standard library is insufficient and one needs
	external packages. Sometimes there are multiple options and the
	user/developer has to make choices regarding the tools to use.
	Especially for the beginners who are not used to the open-source realm,
	it can be confusing prospect to make good technological choices.

@@

So far, we have discussed in detail about Python. However, there are many
interesting open-source languages that can be used for scientific applications.
All have strengths and weaknesses.

>>>>

In order to attain the research goals in this thesis, a path off the beaten
track was pursued -- i.e. to create a handful of easy to maintain, rigorously
tested, and most-importantly, open-source scientific packages which do not
compromise on performance.
%
This resulted in the creation of FluidDyn, a project to facilitate open-science
in the field of fluid mechanics.
%
The focus of the project was to develop a set of packages to implement a
framework for simulations \citep{fluidsim}, experiments
\citep{augier_fluidimage_2016,augier_fluidlab_nodate}, Fast Fourier Transforms
\citep{fluidfft} -- all depending on a base package to reuse certain utilities
and data structures \citep{fluiddyn}.
%
The project owes a lot to the recent developments in methods and tools for
open-source software engineering.
%
Several modern coding practices were adopted while developing these packages.

The FluidDyn packages are primarily written in Python language. Incidentally,
it is the defining characteristics of Python which make the eventual success of
such a project realizable.
%
Python is one of the most important tools in recent open-source dynamics,
particularly in science.
%
The popularity of the language in fluid mechanics is growing, but Python is yet
to reach widespread adoption as the language of choice.

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline, have flourished in the latter half of the twentieth century.
Incidentally, the origin of this discipline can be traced back to the ideas in
\citet{richardson_weather_1922} to achieve numerical weather prediction.
Following the arrival of first electronic general-purpose computer, ENIAC in
1945, the first simulations were made by \citet{charney_numerical_1950} using a
simple barotropic model to make 24 hour forecasts \citep[see
also][]{lynch_richardson_2010}. In those early days when the availability of
computers was limited, such methods were neither widespread nor feasible.
Scientific investigations were dominated by theoretical and experimental
methods.
%
We have come a long way and today computational methods complement theoretical
and experimental studies, and are equally important.
%
% Thus, in a typical scientific study, there is significant effort involved in
% translating the mathematical representation to a working research software.
%
With the connectivity provided by the world wide web it is possible to achieve
more and reinvent the way we do sciences.

@@

In recent years, there has been a move towards open science
\citep[see the report by][]{royal_society_great_britain_science_2012}.
%
Depicted in figure~\ref{fig:opensci} are some of steps needed to ensure
reproducibility and transparency in sciences.
%
Traditionally, \emph{knowledge} has been disseminated in the form of courses
and books from universities, and to the wider scientific community through
research articles, conference proceedings and workshops.
%
In the last decade, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals\footnote{In the European
  Union, a proposal to implement Plan S (\url{https://www.coalition-s.org}) was
  been set in motion in September 2018 to fast track adoption of open-access.
} either through systematic
reforms in academia on how researchers' productivity are assessed, or through
regulations and offering incentives, such as waiving the application processing
charge for publishing \citep{nosek_promoting_2015}.

A lot of work is required to condense and implement this knowledge in the form
of working applications.
%
The \emph{know-how} can be made accessible and open to scrutiny by making
research software open-source with an appropriate license\footnote{For a
	detailed comparison of licenses, visit either
	\url{https://choosealicense.com} or
	\url{https://www.gnu.org/licenses/license-list.html}. }.
%
To make codes usable they have to be complemented by documentation which
typically includes tutorials, examples and a detailed commentary of different
components.
%
Furthermore, to inspire confidence in and to ensure reliability of a code
throughout the process of software development, it needs unit-tests through
which it is continuously built, tested and deployed --- a process known as
Continuous Integration (CI). Depending on how matured a software is, CI can be
easy or cumbersome to implement, but nevertheless, the result is a cleaner and
reliable code.

Finally, the \emph{results} generated from such codes together with the
workflow (i.e.\ scripts to run the codes and to post-process the data) when made
open in the form of datasets can ensure that research is reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

The Free and Open-Source Software (FOSS) movement has given rise to the tools
that enable open-source software development today.
%
The term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success of the type-setting
standards \TeX\ (1977) and \LaTeX\ (1985) for authoring scientific publications.
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development, as described in \citet{raymond_cathedral_1999}, wherein a
code can stay structured despite releasing frequently and delegating tasks to
community with myriad agendas.

@@

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial).
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
In web-based source development platforms like GitHub, Bitbucket and GitLab,
DVCS repositories are enhanced with code-review tools, typically consisting of
a issue tracker and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and low learning barriers for
scientists who are not experts in software development is crucial for our
project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

\subsubsection{Continuous Integration and documentation} The need for
strong reliability of packages with increasing complexity has forced developers
to use Continuous Integration (CI).
%
When the code depends on a wide variety of external dependencies, CI is useful
to avoid \fnref{https://en.wikipedia.org/wiki/Software\_rot}{software rot},
i.e.\ to avoid obsolescence with newer versions of the compiler or its
dependencies. CI is specifically important for dynamical languages and
open-source code since, it allows open-source projects to welcome contributions
from other developers while maintaining a high degree of reliability.
%
Firstly, the code is \emph{built} against a known, repeatable emulation
environment on a server. Thereafter, a set of \emph{unit tests} are run for
each commit (modification of the code) to check that no bugs have been
introduced. It is also important to analyse what portion of the source code is
covered by these unit tests --- a method referred to as \emph{code coverage}.
%
Almost all CI platforms are free for open-source projects --- Travis CI,
CircleCI and Appveyor to name a few. Bitbucket and GitLab have built-in CI
implementations. Websites such as Codecov and Coveralls help to chart and
analyse the code coverage history, viewable by anyone. To use these services,
one simply has to write a simple file in YAML format and authenticate the
respective services with ``read'' permissions.

New users and developers often rely on good documentation and examples.
Documentation generators such as Sphinx (for Python) and Doxygen (for many
languages including C, C++, Fortran and Python) parses in-line comments into
documentations as PDF, HTML and more \citep{lee_ten_2018}.
%
The documentation generation can be automated online and deployed using any CI.
``Read the Docs'' has, in recent years, become the hub for building and hosting
documentation online, thus facilitating the process for developers and
considerably improving access to information for users.
%
In the FluidDyn project we this limit ourselves to automating tests and
documentation of the latest version. It is possible to further exploit the
CI by adding code style checks by running ``linters'' alongside unit tests,
\emph{continuous delivery} of package releases and preserving older versions of
documentation.

@@

% Mutual aid and knowledge sharing have been important aspects of the web since
% its inception and these practices are now of foremost importance for many
% human activities.
%
%In particular, 
Learning programming and computer science is now greatly aided by websites like
the community driven encyclopaedia \href{https://www.wikipedia.org/}{Wikipedia}
or the community driven forum \href{https://stackoverflow.com/}{Stack
Overflow}, where people write thousands of questions and answers on programming
each week. The IRC protocol has continued to sustain the test of time. Instant
messaging channels such as \codeinline{\#python} on Freenode facilitate
interactive discussions and a form of community-level support system --- a role
which used to be filled by mailing lists alone in the past. Alternatives such
as Riot (Matrix protocol), Gitter, and Slack have gained popularity offering
different integration solutions on top of instant messaging.
%
Through instant messaging a geographically separated team, as in the case of
the FluidDyn project, can achieve near real-time collaboration without having
to arrange video conferences or meetings.

Thus, there is a number of new methods and tools in software development.
Taken all together, they greatly improve our collective efficiency and
open possibilities for interactions that were unthinkable a few years ago.
%
For example, the success of community-driven software like \pack{astropy} (a
subject-area research library with
\href{https://github.com/astropy/astropy}{more than 240 contributors}) and
\pack{scikit-learn} (a machine learning toolkit with
\href{https://github.com/scikit-learn/scikit-learn}{more than 1000
contributors}) would not have been possible without these new software
development methods.
%
As scientists, we can wonder how to fully exploit these new possibilities to do
science. The project FluidDyn is an attempt to advance the fluid dynamics
community in this direction.

@@

The Python community has thrived remarkably in the last few years, which is
evident from the fact that by the end of 2017, Python became the second most
popular language in terms of pull-requests in GitHub\footnote{%
%See GitHub's annual report titled \href
\url{https://octoverse.github.com/}
% { ``The State of the Octoverse 2017''}.
%
} and
\fnref%
% [See the blog post ``The Incredible Growth of Python'']%
{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}%
{questions on StackOverflow}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{
%See the analysis titled 
\url{https://spectrum.ieee.org/computing/software/the-2017-top-programming-languages}
%{``The 2017 Top Programming Languages''} on IEEE spectrum.
}
ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings agree that Python usage has been steadily growing and is here to stay.
Python largely outclasses languages like Fortran and Matlab according to these
statistics, and is now clearly one of the mainstream languages with C, Java,
C++ and JavaScript.
%
The language has been designed to boost the communication of technical ideas
between humans. 
%
Let us summarize some characteristics of the Python languages that have lead to
such an incredible success.
%
Python differs from other mainstream languages for a variety of reasons:

@@

\item \textbf{Batteries included.} Python comes with a large
\fnref{https://docs.python.org/3/library/index.html}{standard library}.

@@

\end{itemize}
%
Thus, Python is one of the best ``glue'' languages for fast prototyping.
Code development in Python is much faster and easier than with many other
languages. Less bugs are introduced by the developers simply because there are
fewer lines of code \citep{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also facilitated by the very good code
readability associated with the language.
%
Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy from the start and the
complexity gradually increases as the level of the developer increases.
%
Its interpreted nature is advantageous for a rapid development cycle and the
possibility of interactive workflow (usually with IPython or Jupyter
notebooks).
%
Therefore, Python is good for developers at all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
%
Note that there are many implementations of Python
interpreters\footnote{Notable ones are CPython (written in C), Jython (Java),
IronPython (C\#), PyPy (Rpython, a subset of Python) and
MicroPython/CircuitPython (C, targeted to micro-controllers).}. The default and
most widely used implementation is written in C and is called CPython.
%
A hard and controversial decision was taken to clean up Python from its
historical incoherences by introducing a new series of backwards-incompatible
versions of the language, starting with Python 3.0, released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
completed for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible as of 2018.}. Especially for scientific applications,
we can now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
versions of \Numpy} and \fnref{http://www.python3statement.org/}{the main
scientific packages} will not be compatible with Python 2. These changes
open doors to a very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science any more and
instead, adopt the newer versions.
%
New versions of Python have nice features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).

@@

\end{itemize}
%
Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established
foundational packages
%
(\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There are user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{https://pyzo.org}{Pyzo},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

Python has a very large and supportive community (see, for example,
\fnref{http://stackoverflow.com/tags/python}{Stack Overflow tags}).
%
Python development is also supported by companies, and most of them specialize
in web technologies and data-driven research. However, recently, companies
involved in scientific software development have started to become quite
influential. For example, Anaconda Inc.\ is strongly engaged in the development
of tools like JIT Python accelerator \pack{Numba}, and parallelism framework
\pack{dask}. Hardware giants such as Intel and Nvidia also contribute to the
Python scientific ecosystem.
%
To summarize, Python's many features and its adoption in several domains
makes it a sensible language choice for scientific purposes.

@@

\item \textbf{Concurrent, but not parallel multi-threading.} Threads behave
	like light sub-programs. In principle, they can use the different cores
	of the CPU at the same time, but the CPython interpreter forbid threads
	to interpret python code at the same time\footnote{See
	\url{https://opensource.com/article/17/4/grok-gil}
	%{a blog article ``How to grok the GIL''}
	% and \url{https://faster-cpython.readthedocs.io/cpython37.html}{a
	% technical presentation by a CPython core developer}.
	}. In other words, threads can coexist concurrently, by not execute in
	parallel.
%
This limitation comes from an important detail in the implementation of the
CPython interpreter: a Global Interpreter Lock (GIL) to prevent race conditions
that could corrupt data.
%
GIL greatly simplifies the implementation of CPython and that
it is very difficult to remove it while keeping other nice technical properties
of CPython intact\footnote{See
\url{https://wiki.python.org/moin/GlobalInterpreterLock}. Note that PyPy
also uses a GIL, while two other implementations of the Python language, Jython
and IronPython, do not have this limitation}.

\item \textbf{Lively, and thus, complicated ecosystem of packages.} For many
	applications, the standard library is insufficient and external
	packages are needed. Sometimes there are multiple options and the
	user/developer has to make choices regarding the tools to use.
	Especially for the beginners who are not used to the open-source realm,
	it can be confusing prospect to make good technological choices; but
	this can be solved through proper guidance and training.
	Of course, such an abundance of projects can also be seen as an advantage.

@@

So far, we have discussed the details of Python. However, there are many
interesting open-source languages that can be used for scientific applications.
All have strengths and weaknesses.
